---
title: "Density-Based Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density-Based Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```


## Setup

```{r}
library(workflows)
library(parsnip)
```

Load libraries:

```{r setup}
library(tidyclust)
library(tidyverse)
library(tidymodels)
```


```{r setup_secret, echo = FALSE}
library(ggforce)
set.seed(822)
```

Load and clean a dataset:


```{r}
data("penguins", package = "modeldata")

penguins <- penguins %>%
  select(bill_length_mm, bill_depth_mm) %>%
  drop_na()

# shuffle rows
penguins <- penguins %>%
  sample_n(nrow(penguins))
```

At the end of this vignette, you will find a brief overview of the DBSCAN algorithm.



## `db_clust` specification in {`tidyclust`}

To specify a DBSCAN model in `tidyclust`, simply set the values for `radius` and `min_pts`:

```{r}
db_clust_spec <- db_clust(radius = 0.35, min_points = 10)

db_clust_spec
```

There is currently one engine: `dbscan::dbscan` (default)

## Fitting db_clust models

After specifying the model specification, we fit the model to data in the usual way:

Note that I have standardized bill length and bill depth since DBSCAN uses euclidean distance to fit clusters and these variables are on different scales.

```{r}

penguins_std <- penguins %>%
  mutate(bill_length_std = scale(bill_length_mm)[,1],
         bill_depth_std  = scale(bill_depth_mm)[,1]) %>%
  select(bill_length_std, bill_depth_std)

db_clust_fit <- db_clust_spec %>%
  fit( ~ bill_length_std + bill_depth_std,
       data = penguins_std
      )

db_clust_fit %>% 
  summary()

```

We can also extract the standard `tidyclust` summary list:

```{r}
db_clust_summary <- db_clust_fit %>% extract_fit_summary()

db_clust_summary %>% str()

```

## Cluster assignments and centers

The cluster assignments for the training data can be accessed using the `extract_cluster_assignment()` function.

Note that the DBSCAN algorithm allows for some points to not be assigned clusters. These points are labeled as outliers.


```{r}
db_clust_fit %>% extract_cluster_assignment()
```

### Centroids

While the centroids produced by a db_clust fit may not be of primary interest, they can still be accessed via the `extract_fit_summary()` object.

```{r}
db_clust_summary$centroids
```

## Prediction


Since $DBSCAN$ algorithm ultimately assigns training observations to the clusters based on proximity to core points, it is natural to "predict" that test  observations belong to a cluster if they are within `radius` distance to a core point. To reconcile points being within the radius of two core points that belong to different clusters, the `predict()` function assigns new observations to the cluster of the closest core point. If a point is not within the `radius` of any core points, it is predicted to be an outlier.


```{r}
new_penguin <- tibble(
  bill_length_std = -1.2454235,
  bill_depth_std = 0.5012132
)

db_clust_fit %>%
  predict(new_penguin)
```


## A brief introduction to density-based clustering

Density-Based Spatial Clustering of Applications with Noise (DBSCAN) is a method of unsupervised learning that groups observations into clusters based on their density in multidimensional space. Unlike methods such as k-means, DBSCAN does not require specifying the number of clusters beforehand and can identify clusters of arbitrary shapes. Additionally, it can classify points that do not belong to any cluster as outliers, allowing for greater flexibility in handling real-world data.

In DBSCAN, observations are considered as locations in multidimensional space. The algorithm works by defining a cluster as a dense region of connected points. During the fitting process, points are classified as core points, border points, or noise based on their proximity to other points (**radius**) and a density threshold (**min_points**).

The fitting process can be described as follows:

```{r, echo = FALSE, include = FALSE}

penguins_std %>%
ggplot() +
geom_point(aes(x = bill_length_std, y = bill_depth_std)) +
theme_minimal() +
coord_fixed()

```

1. For each observation in the data, determine whether each point is a core point. A core point is defined as an observation which has more than the specified **min_points** points (including the point itself) within the specified **radius** around the point.


```{r, include = FALSE}
library(dbscan)
eps <- 0.35
min_points <- 10
dbscan_fit <- dbscan(penguins_std, eps = eps, minPts = min_points)

```

```{r, echo = FALSE}
par(mfrow=c(1,2))

penguins_std %>%
  ggplot() +
  geom_point(aes(x = bill_length_std, y = bill_depth_std)) +
  theme_minimal() +
  geom_circle(tibble(x = c(-0.79162259), y = c(0.2), radius = rep(eps,1)),
              mapping = aes(x0 = x, y0 = y, r = eps),
              color = "gray", linewidth = 0.8) +
  geom_point(tibble(x = c(-0.79162259), y = c(0.2)), mapping = aes(x = x, y = y, color = c("1"))) +
  coord_fixed() +
  theme(legend.position = "none") +
  scale_color_manual(values = c("red"))

penguins_new <- penguins_std
penguins_new$cluster <- (db_clust_fit %>% extract_cluster_assignment())$.cluster
# factor(dbscan_fit$cluster)
penguins_new$cp <- factor(if_else(as.numeric(is.corepoint(penguins_std, eps = eps, minPts = min_points)) == 1, "Yes", "No"), levels = c("Yes", "No"))

penguins_new %>%
  ggplot(aes(x = bill_length_std, y = bill_depth_std, color = cp)) +
  geom_point() +
  theme_minimal() +
  coord_fixed() +
  scale_color_manual(values = c("red", "black")) +
  theme(legend.position = "none")
```


2. For each core point, assign it to a cluster as well as all other core points within the specified **radius** of the point. Do until all core points have been assigned to a cluster.

```{r, echo = FALSE}
penguins_new %>%
  mutate(cp_color = if_else(cp == "Yes", cluster, "0")) %>%
  ggplot() +
  geom_point(aes(x = bill_length_std, y = bill_depth_std, color = cp_color)) +
  theme_minimal() +
  coord_fixed() + 
  scale_color_manual(values = c("black", "#F8766D", "#00BA38", "#619CFF")) +
  theme(legend.position = "none")
```

3. For the remaining points not assigned to a cluster, check whether the point is within the radius of a core point. For points not in the radius of any core points, assign the point as an outlier. Otherwise, assign the point to the cluster of the closest core point.


```{r, echo = FALSE}
penguins_new %>%
  ggplot() +
  geom_point(aes(x = bill_length_std, y = bill_depth_std, color = cluster)) +
  theme_minimal() +
  coord_fixed() + 
  scale_color_manual(values = c("#F8766D", "#00BA38", "#619CFF", "black")) +
  theme(legend.position = "none")
```

