---
title: "Gaussian Model Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian Model Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Setup
```{r}
library(workflows)
library(parsnip)
library(mclust)
```

```{r}
library(tidyclust)
library(tidyverse)
library(tidymodels)
set.seed(838383)
```


Load and clean a dataset:

```{r}
data("penguins", package = "modeldata")

penguins <- penguins %>%
  select(bill_length_mm, bill_depth_mm) %>%
  drop_na()


# shuffle rows
penguins <- penguins %>%
  sample_n(nrow(penguins))
```



At the end of this vignette, you will find a brief overview of the GMM algorithm.



## gm clust specification in {`tidyclust`}

```{r}
gm_clust_spec <- gm_clust(num_clusters = 3)

gm_clust_spec
```

There is currently one engine: `mclust::Mclust` (default) 

## Fitting gm_clust models

We fit the model to data in the usual way:

```{r}
gm_clust_fit <- gm_clust_spec %>%
  fit( ~ bill_length_mm + bill_depth_mm,
       data = penguins
      )

gm_clust_fit %>% 
  summary()
```
We can also extract the standard `tidyclust` summary list:


```{r}
# gm_clust_summary <- gm_clust_fit %>% extract_fit_summary()

# gm_clust_summary %>% str()
```


```{r}
gm_clust_fit %>% extract_cluster_assignment()



```

If you have not yet read the `k_means` vignette, we recommend reading that first;
functions that are used in this vignette are explained in more detail there.


## Prediction

<!-- Since unsupervised methods are not designed for prediction, the notion of  -->
<!-- predicting the cluster assignment is not always obvious. -->

write stuff here

```{r}
new_penguin <- tibble(
  bill_length_mm = 40,
  bill_depth_mm = 20
)

gm_clust_fit %>%
  predict(new_penguin)
```


## A brief introduction to guassian mixture models


write and explain here











